\documentclass[a4paper,11pt,article,oneside]{memoir}
\input{../preamble/packages}

\input{../utils/eurodate.tex}
\input{../utils/lingstyles.tex}

\lingset{lingstyle=default}

\setmainfont{Charis SIL}[CharacterVariant=43:1]

\setlength{\parindent}{0em}
\setlength{\parskip}{1ex}
%\linespread{1.1}

\input{../preamble/macros.tex}
\input{../utils/ipa_macros.tex}
\input{../utils/misc_macros.tex}

\title{\langname{} Sound Changes:\\{\Large A Cheat Sheet}}
\author{Bethany E. Toma}
\date{\today}

\counterwithout{section}{chapter}
\maxsecnumdepth{section}

\begin{document}

\maketitle

While \langname{} is distinct enough from its Esperanto ancestor to be reasonably considered a separate language at this juncture, its orthography does not reflect this. While syntactic and semantic changes are reflected to a fairly reasonable extent in the orthography, phonological changes are not accounted for. As a reult, to reasonably approximate the pronunciation of a \langname{} word from how it is written, one must understand which sound changes have occurred from the `original' (hereafter `Zamenhofian') Esperanto. 

Note that some sound changes listed apply only to the Lowlands dialect but not to the Highlands dialect, or vice-versa. This is indicated when it is the case.

\section*{Zamenhofian Esperanto Orthography}

The Esperanto pronunciation of a word tends to be roughly equivalent to its spelling in IPA, with the following exceptions:
\begin{center}
\begin{threeparttable}
\begin{tabular}{ccccccccc}
    \ortho{c} & \ortho{\^{c}} & \ortho{\^{g}} & \ortho{\^{h}}\tnote{1} & \ortho{\^{j}} & \ortho{\^{s}} & \ortho{r} & \ortho{\u{u}} & \ortho{v}\\[0.2cm]
    \phipa{t\tiebar s} & \phipa{t\tiebar\esh} & \phipa{d\tiebar\ezh} & \phipa{x} & \phipa{\ezh} & \phipa{\esh} & \phipa{\alvrap} & \phipa{w} & \phipa{\labrox}
\end{tabular}
\begin{tablenotes}
    \item[1] This phoneme is only retained in a very few words at this point (e.g., \espq{eĥo}, \espq{ĥaoso}) and is elsewhere generally replaced with variants with k or h.
\end{tablenotes}
\end{threeparttable}
\end{center}

Primary stress falls on the penultimate syllable, and secondary stress \emph{tends} to fall on alternating syllables preceding it.

\section{Sound Changes}

\subsection{0--100 AMC}

\input{sound_change_stages/amc_000.tex}

\subsection{100--200 AMC}

\input{sound_change_stages/amc_100.tex}

\subsection{200--300 AMC}

\input{sound_change_stages/amc_200.tex}

\subsection{300--400 AMC}

\input{sound_change_stages/amc_300.tex}

\subsection{400--500 AMC}

\input{sound_change_stages/amc_400.tex}

\subsection{500--600 AMC}

\input{sound_change_stages/amc_500.tex}

\subsection{600--700 AMC}

\input{sound_change_stages/amc_600.tex}

\subsection{700--800 AMC}

\input{sound_change_stages/amc_700.tex}

\subsection{800--900 AMC}

\input{sound_change_stages/amc_800.tex}

\subsection{900--1000 AMC}

\input{sound_change_stages/amc_900.tex}

\section{Sparky's Transcription}

I'll probably make an in-world orthography later, but regardless of whether I do, I need a roughly phonetic or phonemic transcription which I can use to not have to constantly write in IPA. That'll go here once I actually make it lol.

\clearpage
\section{Pronunciation Example}

\input{lords_prayer.tex}

\end{document}